---
title: "Meteorites"
format: html
editor: visual
---

# Set-Up

```{r}
version
```

## Libraries

```{r}
library(dplyr)
library(ggplot2)
library(maps)
library(sf)
library(sp)
library(rworldmap)
library(rnaturalearth)
library(brms)
library(patchwork)
library(ggraph)
library(igraph)
library(RColorBrewer)

```

## Load Data and Create Data Frames

```{r}
space_rocks<-readr::read_csv("~/Desktop/Meteorites/Meteorite_Landings_20250108.csv")

#remove entries with years after 2016 or before 860, which are known errors in the data base
space_rocks<-space_rocks|>subset(year>860 & year <2016)

#fit column name
space_rocks <- space_rocks |> rename(mass_g = `mass (g)`)

#generally rocks with (0,0) coordinates were found somewhere in Antarctica and exact coordinates were not recorded, create flag to note the antarctic ones
space_rocks<-space_rocks|> mutate (zeros = case_when(reclat==0 & reclong==0 ~ TRUE,
                                            reclat!=0 & reclong!=0 ~ FALSE))

#filter out one value with wonky lat
space_rocks<-space_rocks |> filter (reclong <181)

```

# Data Exploration

## World Map

reference:

<https://sarahpenir.github.io/r/making-maps/>

```{r}
#fiter out antartcic because they'll all look like they're in the water (0,0)
world_map_df<-space_rocks|>filter(zeros==FALSE)

world<-map_data("world")

plain <- theme(
  axis.text = element_blank(),
  axis.line = element_blank(),
  axis.ticks = element_blank(),
  panel.border = element_blank(),
  panel.grid = element_blank(),
  axis.title = element_blank(),
  panel.background = element_rect(fill = "white"),
  plot.title = element_text(hjust = 0.5)
)

worldplot <- ggplot() +
  geom_polygon(data = world, aes(x=long, y = lat, group = group), fill="#118ab2") + 
  coord_fixed(1.3)+
  theme(plot.margin = margin(0, 0, 0, 0))+
  plain+
  geom_point(data=world_map_df, 
             aes(x=reclong, y=reclat), 
             size=0.25, colour="#073b4c")+
  labs(title="Locations of Found Meteorites")



worldplot
ggsave("Images/worldplot.png")

```

Most of the relics seem to be in Sweden and quite recent, some sort of samping difference

The rainforest and northern Canada and Russia don't have a lot of finds, no surprise there–fewer people mean fewer finds. The areas in Africa also line up with population density. Again, that makes sense.

## Investigate Meteorite Types

<https://meteoritelab.com/about/meteorites/meteorite-classification/>

```{r}

#all chrondrites are undifferenciated


#extra level of chrondrite sub-division
chondrites<-c("Charbonaceous Chondrites", 
              "Ordinary Chondrites", 
              "Rhumuruti Chondrites", 
              "Estatite Chondrites")


#groups of chondrites and their sub-types
carbonaceous_chondrites<-c("CI", "CM", "CO", "CV", "CK", "CR", "CH", "C")
ordinary_chondrites<-c("H", "L", "LL")
rhumuruti_chondrites<-c() #no code
estatite_chondrites<-c("EH", "EL")

#martian sub-types
martian<-c("shergottite", "nakhlite", "Martian", "Shergottite", "Nakhlite", "chassignite ")

#groups of differentiated meteorites and their subtypes
iron<-c("I", "II", "III", "IV")
achrondrites<-c("Martian", "Aubrites", "Urielites", "Angrites", "HED", "Lunar", "Achondrites", "Acapulcoites", "Winonaites", "Lodranites")
stony<-c("R")
stony_iron<-c("Pallasites", "Mesosiderites")
```

```{r}
#first convert messy column into cleaner types
space_rocks <- space_rocks |>
  mutate(Type = case_when(
    substr(recclass, 1, 2) == "EH" ~ "EH",
    substr(recclass, 1, 2) == "EL" ~ "EL",
    
    substr(recclass, 1, 2) == "LL" ~ "LL",
    grepl("^H[0-9]", substr(recclass, 1, 2)) ~ "H",
    substr(recclass, 1, 2) == "H/" ~ "H",
    substr(recclass, 1, 2) == "H-" ~ "H",
    substr(recclass, 1, 2) == "H(" ~ "H",
    substr(recclass, 1, 2) == "H~" ~ "H",
    grepl("^H$", recclass) ~ "H", 
    grepl("^L[0-9]", substr(recclass, 1, 2)) ~ "L",
    grepl("^L$", recclass) ~ "L", 
    substr(recclass, 1, 2) == "L/" ~ "L",
    substr(recclass, 1, 2) == "L(" ~ "L",
    substr(recclass, 1, 2) == "L~" ~ "L",
    substr(recclass, 1, 2) == "L-" ~ "L",
    
    substr(recclass, 1, 2) == "CI" ~ "CI",
    substr(recclass, 1, 2) == "CM" ~ "CM",
    substr(recclass, 1, 2) == "CO" ~ "CO",
    substr(recclass, 1, 2) == "CV" ~ "CV",
    substr(recclass, 1, 2) == "CK" ~ "CK",
    substr(recclass, 1, 2) == "CR" ~ "CR",
    substr(recclass, 1, 2) == "CH" ~ "CH",
    grepl("^C[0-9]", substr(recclass, 1, 2)) ~ "C",
    
    substr(recclass, 1, 3) == "How" ~ "Howardites",
    substr(recclass, 1, 3) == "Euc" ~ "Eucrites",
    substr(recclass, 1, 3) == "Dio" ~ "Diogenites",
    
    substr(recclass, 1, 3) == "Ach" ~ "Achondrites",
    substr(recclass, 1, 3) == "Aca" ~ "Acapulcoites",
    substr(recclass, 1, 3) == "Win" ~ "Winonaites",
    substr(recclass, 1, 3) == "Lod" ~ "Lodranites",
    
    substr(recclass, 1, 3) == "Pal" ~ "Pallasites",
    substr(recclass, 1, 3) == "Mes" ~ "Mesosiderites",
    
    substr(recclass, 1, 3) == "ALH" ~ "ALH 84001",
    
    substr(recclass, 1, 3) == "Aub" ~ "Aubrites",
    substr(recclass, 1, 3) == "Ure" ~ "Ureilites",
    substr(recclass, 1, 3) == "Ang" ~ "Angrites",
    
    substr(recclass, 1, 1) == "R" ~ "R",
    
    substr(recclass, 1, 3) == "Lun" ~ "Lunar",
    
    grepl("I{1,4}", recclass) ~ sub(".*?(I{1,4}).*", "\\1", recclass),
    
    substr(recclass, 1, 3) == "Unk" ~ "Unknown",
    
    grepl("^Martian$", recclass) ~ "Martian",
    grepl("^Martian \\(.*\\)$", recclass) ~ sub("^Martian \\((.*)\\)$", "\\1", recclass),
    
    substr(recclass, 1, 3) == "she" ~ "Shergottites",
    substr(recclass, 1, 3) == "nah" ~ "Nakhlites",
    substr(recclass, 1, 3) == "cha" ~ "Chassigns",
    
    substr(recclass, 1, 3) == "She" ~ "Shergottites",
    substr(recclass, 1, 3) == "Nah" ~ "Nakhlites",
    substr(recclass, 1, 3) == "Cha" ~ "Chassigny",
    
    TRUE ~ "Other"
  ))


space_rocks$martian <- ifelse(space_rocks$Type %in% martian, "martian", NA)
```

```{r}
#add labels for undifferentiated meteorites
space_rocks <- space_rocks |>
  mutate(chondrites = case_when(
    Type %in% carbonaceous_chondrites ~ "carbonaceous_chondnrites",
    Type %in% ordinary_chondrites ~ "ordinary_chrondrites",
    Type %in% rhumuruti_chondrites ~ "rhumuruti_chrondrites",
    Type %in% estatite_chondrites ~ "estatite_chrondrites",
    TRUE ~ NA
           )
  )

#add labels for differentiated meteorites
space_rocks <- space_rocks |>
  mutate(achondrites = case_when(
    Type %in% achrondrites ~ "achrondrites",
    Type %in% iron ~ "iron",
    Type %in% stony ~ "stony",
    Type %in% stony_iron ~ "stony_iron",
    TRUE ~ NA
           )
  )

#add labels for top category
space_rocks <- space_rocks |>
  mutate(differentiated = case_when(
    Type %in% carbonaceous_chondrites ~ "undifferentiated",
    Type %in% ordinary_chondrites ~ "undifferentiated",
    Type %in% rhumuruti_chondrites ~ "undifferentiated",
    Type %in% estatite_chondrites ~ "undifferentiated",
    Type %in% achondrites ~ "differentiated",
    Type %in% iron ~ "differentiated",
    Type %in% stony ~ "differentiated",
    Type %in% stony_iron ~ "differentiated",
    TRUE ~ NA
           )
  )


```

## Mass: Differentiated vs Undifferentiated

```{r}
#| echo: false


density_group_plot<-space_rocks|> filter (!is.na(mass_g) & !is.na(differentiated) )|>
  ggplot(aes(x=mass_g, colour=differentiated))+
  geom_density()+
  scale_x_log10()+
  ggtitle("Mass & Differentiation Status")+
  labs(x="Mass in Grams [Log Scale]", y="Density", colour="Melted?")+
  scale_color_manual(values=c("#ef476f", "#06d6a0"), labels=c("Differentiated", "Undifferentiated"))+
  theme(legend.position = "bottom", 
        legend.box = "horizontal")

density_group_plot
```

## Mass: By Meteorite Type

```{r}

density_plot_df<-space_rocks|> filter (!is.na(mass_g) & !is.na(differentiated))

density_plot_df$Type <- gsub("Mesosiderites", "Mes", density_plot_df$Type)
density_plot_df$Type <- gsub("Pallasites", "Pal", density_plot_df$Type)

density_group_plot_all<- density_plot_df|>
  ggplot(aes(x=mass_g, colour=Type))+
  geom_density()+
  scale_x_log10()+
  ggtitle("Mass & Meteorite Type")+
  labs(x="Mass in Grams [Log Scale]", y="Density")+
  theme(legend.position = "bottom", 
        legend.box = "horizontal")


density_group_plot_all
```

## Mass: Differentiation and Type

```{r}
mass_density_plots<-density_group_plot + density_group_plot_all

mass_density_plots
ggsave("Images/mass_density_plots.png")
```

## Prevalence of Meteorites by Type

```{r}

type_count_df<- space_rocks |>filter (Type != "Unknown" & Type != "Other") |> group_by(Type) |> summarise (count=n())

Number_of_each_type<-type_count_df|>ggplot(aes(x=Type, y=count, colour=Type, fill=Type))+
 geom_bar(stat="identity")+
  labs(title="Prevalence of Meteorites by Type", x= "Type", y="Count")+
  theme(legend.position = "bottom", 
        legend.box = "horizontal")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
Number_of_each_type
ggsave("Images/Number_of_each_type.png")
```

## Exploring Other Differences in Mass

```{r}
density_group_plot_chondrites<-space_rocks|> filter (!is.na(mass_g) & !is.na(chondrites) )|>
  ggplot(aes(x=mass_g, colour=chondrites))+
  geom_density()+
  scale_x_log10()+
  ggtitle("Differences in Mass by Meteorite Type")

density_group_plot_chondrites

density_group_plot_achondrites<-space_rocks|> filter (!is.na(mass_g) & !is.na(achondrites) )|>
  ggplot(aes(x=mass_g, colour=achondrites))+
  geom_density()+
  scale_x_log10()+
  ggtitle("Differences in Mass by Meteorite Sub-Type")

density_group_plot_achondrites


density_group_plot_martian<-space_rocks|> filter (!is.na(mass_g) 
                                                  & !is.na(martian) 
                                                  & Type != "Martian")|>
  ggplot(aes(x=mass_g, colour=Type))+
  geom_density()+
  scale_x_log10()+
  ggtitle("Differences in Mass by Meteorite Type")

density_group_plot_martian
```

## Boxplots: Mass with Different Categories

```{r}

boxplot_group_plot_chondrites<-space_rocks|> filter (!is.na(mass_g) & !is.na(chondrites))|>
  ggplot(aes(x=differentiated, y=mass_g, colour=chondrites))+
  geom_boxplot()+
  ggtitle("Differences in Mass by Meteorite Type")

boxplot_group_plot_chondrites

space_rocks_less1000g<-space_rocks|>filter(mass_g<1000)


boxplot_group_plot_chondrites<-space_rocks|> filter (!is.na(mass_g) & !is.na(chondrites))|>
  ggplot(aes(x=differentiated, y=mass_g, colour=chondrites))+
  geom_boxplot()+
  ggtitle("Differences in Mass by Meteorite Type")

boxplot_group_plot_chondrites

space_rocks_less1000g<-space_rocks|>filter(mass_g<1000)



boxplot_group_plot_chondrites_smaller<-space_rocks_less1000g|> filter (!is.na(mass_g) & !is.na(chondrites))|>
  ggplot(aes(x=differentiated, y=mass_g, colour=chondrites))+
  geom_boxplot()+
  ggtitle("Differences in Mass by Meteorite Type")

boxplot_group_plot_chondrites_smaller

boxplot_group_plot_martian_smaller<-space_rocks_less1000g|> filter (!is.na(mass_g) &
                                                                    !is.na(differentiated))|>
  ggplot(aes(x=differentiated, y=mass_g, colour=differentiated))+
  geom_boxplot()+
  ggtitle("Differences in Mass by Meteorite Type: Mars")

boxplot_group_plot_martian_smaller



moon_vs_mars<-space_rocks|>filter(martian=="martian" | Type=="Lunar")
moon_vs_mars<- moon_vs_mars|>mutate (martian = case_when(martian=="martian" ~ "martian",
                                            is.na(martian) ~ "Lunar"))


mars_vs_moon_boxplot<-moon_vs_mars|> filter (!is.na(mass_g))|>
  ggplot(aes(x=martian, y=mass_g, colour=martian))+
  geom_boxplot()+
  ggtitle("Differences in Mass by Meteorite Type: Moon Vs Mars")

mars_vs_moon_boxplot



```

## Mass: Lunar vs. Martian Meteorites

```{r}
#moon vs mars density plot

density_group_plot_moon_vs_mars<-moon_vs_mars|> filter (!is.na(mass_g))|>
  ggplot(aes(x=mass_g, colour=martian))+
  geom_density()+
  ggtitle("Differences in Mass by Meteorite Type")

density_group_plot_moon_vs_mars

density_group_plot_moon_vs_mars_log10<-moon_vs_mars|> filter (!is.na(mass_g))|>
  ggplot(aes(x=mass_g, colour=martian))+
  geom_density()+
  scale_x_log10()+
  ggtitle("Differences in Mass by Meteorite Type")

density_group_plot_moon_vs_mars_log10
```

## Year of Discovery vs. Mass

```{r}
time_vs_mass<-space_rocks|>ggplot(aes(x=year, y=mass_g, colour=Type))+
  geom_point()

time_vs_mass
```

# Hierarchical Bayesian Linear Regression

## Data Prep

```{r}
#prep data for Hierachial Bayesian Linear Regression

#want 5-10 per group, check group counts

#create data frame to count rock types occurances in original DF
type_count_df<-data.frame(rock_types=space_rocks$Type|>unique(),
                          sample_count=NA)

type_count_df <- type_count_df |> mutate (sample_count = 
                                            sapply(rock_types, function(x) sum(space_rocks$Type == x)))

#flag anything with 10 or fewer samples
type_count_df<- type_count_df |> mutate (include_flag = ifelse (sample_count>=10, 1, 0))

#make list ofrock_type to include
include_rocks<-type_count_df|>filter(include_flag==1)|>pull(rock_types)

#filter out ones that aren't in the list
bayes_prep_df<-space_rocks|>filter (Type %in% include_rocks)

#recode things nicely
bayes_prep_df <- bayes_prep_df |> mutate (group =  case_when(
    Type %in% carbonaceous_chondrites ~ "chondrites",
    Type %in% ordinary_chondrites ~ "chondrites",
    Type %in% rhumuruti_chondrites ~ "chondrites",
    Type %in% estatite_chondrites ~ "chondrites",
    Type %in% achrondrites ~ "achrondrites",
    Type %in% iron ~ "iron",
    Type %in% stony ~ "stony",
    Type %in% stony_iron ~ "stony_iron",
    TRUE ~ NA
           )
  )

#filter out NA values in differentiated
bayes_prep_df <- bayes_prep_df |> filter (!is.na(differentiated ))

#get DF ready
#columns we want
bayes_ready_df<- bayes_prep_df |> select ("name", "id", "mass_g", "Type", "group", "differentiated")

#make sure there are no NAs
bayes_ready_df <- na.omit(bayes_ready_df)

bayes_ready_df <- bayes_ready_df |> mutate (across (c("Type", "group", "differentiated"), as.factor))

#filter out ones that have zero as mass
bayes_ready_df<- bayes_ready_df |>filter(mass_g !=0)
```

## Run Initial Model

```{r}
bayes_space_rock_model <-brm(
  formula = mass_g ~ 1 + (1 | Type) + (1 + group) + (1 | differentiated), #fixed intercepts, all 3 factors as random effects
  data = bayes_ready_df,
  prior = c(
    prior (normal (0,5), class = "b"), #fixed effects/the intercept
    prior (normal (0, 5), class = "sd") #random effects
  ),
  chains = 4,
  iter = 2000,
  warmup =1000
)
```

Note: can use control = list(adapt_delta = 0.95) if problems with convergence

Model takes considerable time to run: save/load model as appopriate

```{r}
#save model
#saveRDS(bayes_space_rock_model, "~/Desktop/space_rocks_model")

bayes_space_rock_model<-readRDS("~/Desktop/Meteorites/space_rocks_model")


```

## Examine Initial Model

```{r}
#summary
summary(bayes_space_rock_model)

#post. samples
posterior_samples(bayes_space_rock_model)
```

Notes:

-   It converged, but there's a lot of uncertainty. Try again but with a log transform like in the charts.

-   Log transform gets stuck immediately. Add scalar coefficient to try.

-   Still gets stuck, change acceptance rate and try again

-   Still gets stuck, try to get rid of middle level?

-   didn't work, get rid of top level, put middle level back

## Run Second Model 

Note: models that did not work were discarded

```{r}


bayes_space_rock_model2 <-brm(
  formula = 2*log10(mass_g) ~ 1 + (1 | Type) + (1 + group) , #fixed intercepts, all 3 factors as random effects
  data = bayes_ready_df,
  prior = c(
    prior (normal (0,5), class = "b"), #fixed effects/the intercept
    prior (normal (0, 5), class = "sd") #random effects
  ),
  chains = 4,
  iter = 2000,
  warmup =1000,
  control = list(adapt_delta = 0.99)
)
```

```{r}
saveRDS(bayes_space_rock_model2, "~/space_rocks_model2")
```

## Examine Second Model

```{r}

bayes_space_rock_model2<-readRDS("./space_rocks_model2")

fixef(bayes_space_rock_model2)

ranef(bayes_space_rock_model2)

pp_check(bayes_space_rock_model2)
```

Our estimates and C.I.s are much better than before but some are negative.

## Run Third Model

Try log-normal family.

```{r}
bayes_space_rock_model3 <-brm(
  formula = mass_g ~ 1 + (1 | Type) + (1 + group) , #fixed intercepts, all 3 factors as random effects
  data = bayes_ready_df,
  family = lognormal(),
  prior = c(
    prior (normal (0,5), class = "b"), #fixed effects/the intercept
    prior (normal (0, 5), class = "sd") #random effects
  ),
  chains = 4,
  iter = 2000,
  warmup =1000,
  control = list(adapt_delta = 0.99)
)
```

```{r}
saveRDS(bayes_space_rock_model3, "~/space_rocks_model3")
```

```{r}
#summary
summary(bayes_space_rock_model3)

#post. samples
posterior_samples(bayes_space_rock_model3)

pp_check(bayes_space_rock_model3)
#posterior check not good, go with second model
```

```{r}
#pull out group level effects
ranef(bayes_space_rock_model3)
```

# Additional Graphics for Presentation

## Tree Diagram

```{r}

#generate lists of edges in data frames
first_split_from<-c("Meteorites", "Meteorites")
first_split_to<-bayes_ready_df$differentiated|>unique()

myDF1<-data_frame(first_split_from, first_split_to)
myDF1|>colnames()<-c("from", "to")
myDF1$to<-myDF1$to|>as.character()
myDF1[]<-lapply(myDF1, function(x) tools::toTitleCase(x))
myDF1$to<-myDF1$to|>as.factor()

myDF2<-bayes_ready_df|>select("differentiated", "group")|>unique()
myDF2|>colnames()<-c("from", "to")
myDF2[]<-lapply(myDF2, function(x) as.character(x))
myDF2[]<-lapply(myDF2, function(x) tools::toTitleCase(x))
myDF2$to <- gsub("Stony_iron", "SI", myDF2$to)

myDF3<-bayes_ready_df|>select("group", "Type")|>unique()
myDF3|>colnames()<-c("from", "to")
myDF3[]<-lapply(myDF3, function(x) as.character(x))
myDF3[]<-lapply(myDF3, function(x) tools::toTitleCase(x))

myDF3$to <- gsub("Mesosiderites", "Mes", myDF3$to)
myDF3$to <- gsub("Pallasites", "Pal", myDF3$to)
myDF3$from <- gsub("Stony_iron", "SI", myDF3$from)


myEdges<-rbind(myDF1, myDF2, myDF3)
myEdges$shades<-c("A", "B", "A", "C", "D", "E", "H", "I", "J", "K", "C", "l", "M", "N", "F", "P", "Q", "R", "S", "G", "E", "T")

# Create a graph object
myGraph <- graph_from_data_frame(myEdges)


myTree<-ggraph(myGraph, layout = 'dendrogram', circular = FALSE) + 
  geom_edge_diagonal(aes(color = shades)) +
  ggtitle ("Meteorites by Type")+
  scale_colour_identity()+
  geom_node_text(aes( label=name) , angle=0 , hjust=0, nudge_y=-0.1, nudge_x = 0.05) +
  geom_node_point(aes(filter=leaf) , alpha=0.6) +
  ylim(-.5, NA)+
  xlim(NA, 16)+
  theme(legend.position = "none")
  

myTree
ggsave("Images/Tree.png")
```

## Group Vs Type Density

```{r}
density_group_plot_group<-bayes_ready_df|> 
  ggplot(aes(x=mass_g, colour=group))+
  geom_density()+
  scale_x_log10()+
  ggtitle("Mass & Meteorite Group")+
  labs(x="Mass in Grams [Log Scale]", y="Density", colour="group")+
  scale_color_manual(values=c("#073B4C", "#06d6a0", "#FFD166", "#EF476F"), labels=c("Chond.", "Iron", "Stony", "S-I"))+
  theme(legend.position = "bottom", 
        legend.box = "horizontal")

density_group_plot_group
ggsave("Images/density_group_plot_group.png")


mass_density_plots2<-density_group_plot_group + density_group_plot_all

mass_density_plots2
ggsave("Images/mass_density_plots2.png")

```
